Plug-and-play dual-tree algorithm runtime analysis

### 2014-12-24 bounds-files/conclusion.tex
Ready for review.

We have presented a unified framework for bounding the runtimes of dual-tree
algorithms that use cover trees and the standard cover tree pruning dual-tree
traversal (Algorithm \ref{alg:cover-tree-dual}).  Our main result, Theorem
\ref{thm:ct-runtime}, allows plug-and-play runtime bounding of these algorithms.
We have shown that Theorem \ref{thm:ct-runtime} is useful for bounding the
runtime of nearest neighbor search (Theorem \ref{thm:nns}), approximate kernel
density estimation (Theorem \ref{thm:kde-bound}), exact range count, and exact
range search (Theorem \ref{thm:rs}).  With our contribution, bounding a cover
tree dual-tree algorithm is straightforward and only involves bounding the
maximum size of the reference set, $| R^* |$.

>>>>

We have presented a unified framework for bounding the runtimes of dual-tree
algorithms that use cover trees and the standard cover tree pruning dual-tree
traversal (Algorithm \ref{alg:cover-tree-dual}).  In order to produce an
understandable bound, we have introduced the notion of cover tree imbalance; one
possible interesting direction of future work is to empirically and
theoretically minimize this quantity by way of modified tree construction
algorithms; this is likely to provide both tighter runtime bounds and also
accelerated empirical results.

Our main result, Theorem \ref{thm:ct-runtime}, allows plug-and-play runtime
bounding of these algorithms.  We have shown that Theorem \ref{thm:ct-runtime}
is useful for bounding the runtime of nearest neighbor search (Theorem
\ref{thm:nns}), approximate kernel density estimation (Theorem
\ref{thm:kde-bound}), exact range count, and exact range search (Theorem
\ref{thm:rs}).  With our contribution, bounding a cover tree dual-tree algorithm
is straightforward and only involves bounding the maximum size of the reference
set, $| R^* |$.

### 2014-12-24 bounds-files/general-runtime-bound.tex
Ready for review.

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

@@

This dual-tree recursion is a depth-first recursion in the query tree and a
breadth-first recursion in the reference tree.  It descends either the query
tree or the reference tree (not both) during a single iteration, and the
conditional in line \ref{alg:line:ct-dual-ref-recursion-start} ensures that the
combinations of query and reference nodes that are being visited are close in
scale.

Due to the favorable theoretical properties of the cover tree, we can prove
general runtime bounds when this traversal is used.

@@

\begin{equation}
O\left(c_r^4 | R^* | \chi \psi (N + \theta)\right)
\end{equation}

\noindent time, where $ | R^* | $ is the maximum size of the reference set $R$
(line \ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
possible runtime of \texttt{Score()}, and $\theta$ is a measure of imbalance
defined by

\begin{equation}
\theta = \max(\log_2(\eta_r / \eta_q), 0) + i_t(\mathscr{T}_q) + N
\max(\log_2(\delta_q / \delta_r), 0)
\end{equation}

\noindent with $\eta_q$ as the maximum pairwise distance between points in
$S_q$, $\delta_q$ as the minimum nonzero pairwise distance between points in
$S_q$, $\eta_r$ as the maximum pairwise distance between points in $S_r$, and
$\delta_r$ as the minimum nonzero pairwise distance between points in $S_r$.
\end{thm}

@@

First, consider case \textit{(1)}.  If the scale of the root of the query tree
is $s_q^T$ and the scale of the root of the reference tree is $s_r^T$, then
$s_r^T - s_q^T$ recursions of this sort occur before the first query recursion.

@@

Lastly, case \textit{(3)} occurs when the reference scale is less than the
minimum query scale $s_q^{\min}$; that is, all of the query recursions are done.
Define the minimum reference scale as $s_r^{\min}$.  If $s_q^{\min} - s_r^{\min}
> 0$, each query node may have up to $s_q^{\min} - s_r^{\min}$ reference
recursions of type \textit{(3)}, and there are $O(N)$ query nodes.  Thus, the
total number of reference recursions of type \textit{(3)} is bounded by $O(N)
\max(s_q^{\min} - s_r^{\min}, 0)$.

Combining these three cases, we get the following expression for an upper bound
$\theta$ on the total number of reference recursions:

\begin{equation}
\theta = \max(s_r^T - s_q^T, 0) + b(\mathscr{T}_q) + O(N) + O(N) (s_q^{\min} -
s_r^{\min}).
\end{equation}

Now, we can express these in terms of more common constants which are related to
the aspect ratio of the data.  Define $\delta_q$ as the minimum nonzero pairwise
distance in $S_q$ and $\eta_q$ as the furthest pairwise distance in $S_q$ (and
define $\delta_r$ and $\eta_r$ accordingly for the set $S_r$).  Then, $s_r^T =
\lceil \log_2 \eta_r \rceil$, $s_q^T = \lceil \log_2 \eta_q \rceil$, $s_r^{\min}
= \lceil \log_2 \delta_r \rceil$, and $s_q^{\min} = \lceil \log_2 \delta_q
\rceil$.  This gives a more reasonable expression:

\begin{equation}
\theta = \max\left(\log_2\left(\frac{\eta_r}{\eta_q}\right), 0\right) +
b(\mathscr{T}_q) + O(N) + O(N)
\max\left(\log_2\left(\frac{\delta_q}{\delta_r}\right), 0\right).
\end{equation}

Returning to our original line of reasoning, we can combine these results to
give the result of the theorem:

\begin{equation}
O(c_r^4 \psi \chi | R^* | \theta) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
\chi |R^*| (N + \theta)).
\end{equation}
\end{proof}

@@

The bichromatic case ($S_q \ne S_r$) is a slightly more complex result which
deserves a bit more attention.  The intuition for all terms except $\theta$
remain virtually the same, although it is worth pointing out that there is
no dependence on the reference set size.

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has two corner cases: all extra reference recursions are of
type \textit{(1)}, and all extra reference recursions are of type \textit{(3)}.
The first case presents the best situation, because all of the reference
recursions are done before the first query node is recursed into\footnote{In
fact, re-derivation for this specific case can remove the dependence on
$i_t(\mathscr{T}_q)$ entirely.  But our bound, as derived, does not bound this
behavior tightly.  To do so would introduce a number of complex constants, and
thus in that respect $i_t(\mathscr{T}_q)$ is a more suitable bounding quantity}.
Therefore the bulk of the reference recursion work is done by with the query
tree root as the query node; this is the best case.  The worst case is the
latter, where recursion in the reference tree cannot begin until reaching the
bottom of the query tree.  Thus, no pruning can occur until reaching the leaves
of the query tree, and the recursion is essentially equivalent to a single-tree
recursion for each query point plus the added overhead of the recursion of the
query tree with one reference node.

{\bf TODO: what does the second case degenerate to in terms of running time?}

@@

Our results are similar to that of \citet{ram2009}, but those results depend on
a quantity called the {\it constant of bichromaticity}, denoted $\kappa$, which
has unclear relation to cover tree imbalance.  The dependence on $\kappa$ is
given as $c_q^{4 \kappa}$, which is not a good bound, especially because
$\kappa$ may easily be very large.  Further, it is incorrectly claimed that
$\kappa = 1$ in the monochromatic case.

@@

Both $\nu$ and $\kappa$ are difficult to empirically calculate and require an
entire run of the dual-tree algorithm.  On the other hand, bounding
$i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
(assuming the tree is already built).  Thus, not only is our bound tighter, it
more closely reflects the actual behavior of dual-tree algorithms, and the
constants which it depends upon are straightforward to calculate.

>>>>

    \medskip
    \STATE $s^{\max}_r \gets \max_{\mathscr{N}_r \in R} s_r$
\label{alg:line:ct-dual-srmax}
    \IF{$(s_q < s^{\max}_r)$ {\bf or} $(s^{\max}_r = -\infty)$}
\label{alg:line:ct-dual-ref-recursion-start}
      \FORALL{$\mathscr{N}_r \in R$} \label{alg:line:ct-dual-base-case-start}
        \STATE \texttt{BaseCase($p_q$, $p_r$)}
      \ENDFOR \label{alg:line:ct-dual-base-case-end}
      \STATE $R_r \gets \{ \mathscr{N}_r \in R : s_r = s^{\max}_r \}$
\label{alg:line:ct-dual-ref-set}
      \STATE $R_{r - 1} \gets \{ \mathscr{C}(\mathscr{N}_r) : \mathscr{N}_r \in
R_r \} \cup (R \setminus R_r)$ \label{alg:line:ct-dual-ref-children}
      \STATE $R'_{r - 1} \gets \{ \mathscr{N}_r \in R_{r - 1} :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-ref-score}
      \STATE recurse with $\mathscr{N}_q$ and $R'_{r - 1}$
\label{alg:line:ct-dual-ref-recursion-end}
    \ELSE \label{alg:line:ct-dual-query-recursion-start}
      \FORALL{$\mathscr{N}_{qc} \in \mathscr{C}(\mathscr{N}_q)$}
        \STATE $R' \gets \{ \mathscr{N}_r \in R :
\texttt{Score(}\mathscr{N}_q\texttt{,} \mathscr{N}_r\texttt{)} \ne \infty \}$
\label{alg:line:ct-dual-query-pruning}
        \STATE recurse with $\mathscr{N}_{qc}$ and $R'$
\label{alg:line:ct-dual-query-recursion}
      \ENDFOR \label{alg:line:ct-dual-query-recursion-end}
    \ENDIF
  \end{algorithmic}
  \caption{The standard pruning dual-tree traversal for cover trees.}
  \label{alg:cover-tree-dual}
\end{algorithm}

@@

This dual-tree recursion is a depth-first recursion in the query tree and a
breadth-first recursion in the reference tree; to this end, the recursion
maintains one query node $\mathscr{N}_q$ and a reference set $R$.  The set $R$
may contain reference nodes with many different scales; the maximum scale in the
reference set is $s_r^{\max}$ (line \ref{alg:line:ct-dual-srmax}).  Each single
recursion will descend either the query tree or the reference tree, not both;
the conditional in line \ref{alg:line:ct-dual-ref-recursion-start}, which
determines whether the query or reference tree will be recursed, is aimed at
keeping the relative scales of query nodes and reference nodes equal.

A query recursion (lines
\ref{alg:line:ct-dual-query-recursion-start}--\ref{alg:line:ct-dual-query-recursion-end})
is straightforward: for each child $\mathscr{N}_{qc}$ of $\mathscr{N}_q$, the
node combinations $(\mathscr{N}_{qc}, \mathscr{N}_r)$ are scored for each
$\mathscr{N}_r$ in the reference set $R$.  If possible, these combinations are
pruned to form the set $R'$ (line \ref{alg:line:ct-dual-query-recursion}), and
then the algorithm recurses with $\mathscr{N}_{qc}$ and $R'$.

A reference recursion (lines
\ref{alg:line:ct-dual-ref-recursion-start}--\ref{alg:line:ct-dual-ref-recursion-end})
is similar to a query recursion, but the pruning strategy is significantly more
complicated.  We score each node combination $(\mathscr{N}_q, \mathscr{N}_r)$
for all $\mathscr{N}_r$ in $R_{r - 1}$, which is the set of children of all
nodes in $R$ that have scale $s_r^{\max}$ (that set is $R_r$).  The set of
reference nodes in $R_{r - 1}$ that were not pruned create the set $R'_{r - 1}$.
Then, this set is combined with $R \setminus R_r$---that is, each of the nodes
in $R$ that was {\em not} recursed into---to produce $R'$, and the algorithm
recurses with $\mathscr{N}_q$ and the reference set $R'$.

The reference recursion only recurses into the top-level subset of the reference
nodes in order to preserve the separation invariant.  It is easy to show that
every pair of points held in nodes in $R$ is separated by at least
$2^{s_r^{\max}}$:

\begin{lemma}
For all nodes $\mathscr{N}_i, \mathscr{N}_j \in R$ (in the context of Algorithm
\ref{alg:cover-tree-dual}) which contain points $p_i$ and $p_j$, respectively,
$d(p_i, p_j) > 2^{s_r^{\max}}$, with $s_r^{\max}$ defined as in line
\ref{alg:line:ct-dual-srmax}.
\end{lemma}

\begin{proof}
This proof is by induction.  If $|R| = 1$, such as during the first reference
recursion, the result obviously holds.  Now consider any reference set $R$ and
assume the statement of the lemma holds for this set $R$, and define
$s_r^{\max}$ as the maximum scale of any node in $R$.  Construct the set
$R_{r - 1}$ as in line \ref{alg:line:ct-dual-ref-children} of Algorithm
\ref{alg:cover-tree-dual}.

Now, take any $\mathscr{N}_i, \mathscr{N}_j$ in $R_{r - 1}$, with points $p_i$
and $p_j$, respectively, and scales $s_i$ and $s_j$, respectively.  Clearly, if
$s_i = s_j = s_r^{\max} - 1$, then by the separation invariant $d(p_i, p_j) >
2^{s_r^{\max} - 1}$.

Now suppose that $s_i < s_r^{\max} - 1$.  This implies that there exists some
implicit cover tree node with point $p_i$ and scale $s_r^{\max} - 1$ (as well an
implicit child of this node $p_i$ with scale $s_r^{\max} - 2$ and so forth until
one of these implicit nodes has child $p_i$ with scale $s_i$).  Because the
separation invariant applies to both implicit and explicit representations of
the tree, we conclude that $d(p_i, p_r) > 2^{s_r^{\max}} - 1$.  The same
argument may be made for the case where $s_j < s_r^{\max} - 1$, with the same
conclusion.

We may therefore conclude that each point of each node in $R_{r - 1}$ is
separated by $2^{s_r^{\max} - 1}$.  Note that $R'_{r - 1} \subseteq R_{r - 1}$
and that $R \setminus R_{r - 1} \subseteq R$, and that each point of each node
is separated by $2^{s_r^{\max} - 1}$, to see that this condition holds for all
nodes in $R' = R'_{r - 1} \cup (R \setminus R_{r - 1})$.

Because we have shown that the condition holds for the initial reference set and
for any reference set produced by a reference recursion, we have shown that the
statement of the lemma is true.
\end{proof}

This observation means that the set of points $P$ held by all nodes in $R$ is
always a subset of $C_{s_r^{\max}}$.  This fact will be useful in our later
runtime proofs.

Next, we develop notions with which to understand the behavior of the cover tree
dual-tree traversla when the datasets are of significantly different scale
distributions.

If the datasets are similar in scale distribution (that is, inter-point distances
tend to follow the same distribution), then the recursion will alternate between
query recursions and reference recursions.  But if the query set contains points
which are, in general, much further apart than the reference set, then the
recursion will start with many query recursions before reaching a reference
recursion.  The converse case also holds.  We are interested in formalizing
this notion of scale distribution; therefore, define the following
dataset-dependent constants for the query set $S_q$ and the reference set $S_r$:

\begin{itemize}
  \item $\eta_q$: the largest pairwise distance in $S_q$
  \item $\delta_q$: the smallest nonzero pairwise distance in $S_q$
  \item $\eta_r$: the largest pairwise distance in $S_r$
  \item $\delta_r$: the smallest nonzero pairwise distance in $S_r$
\end{itemize}

These constants are directly related to the aspect ratio of the datasets;
indeed, $\eta_q / \delta_q$ is exactly the aspect ratio of $S_q$.  Further,
let us define the top and bottom levels of each tree:

\begin{itemize}
  \item The {\it top scale} of the query tree $\mathscr{T}_q$ is defined as
$s_q^T = \lceil \log_2(\eta_q) \rceil$.
  \item The {\it minimum scale} of the query tree $\mathscr{T}_r$ is defined as
$s_q^{\min} = \lceil \log_2(\delta_q) \rceil$.
  \item The top scale of the reference tree $\mathscr{T}_r$ is defined as $s_r^T
= \lceil \log_2(\eta_r) \rceil$.
  \item The minimum scale of the reference tree $\mathscr{T}_r$ is defined as
$s_r^{\min} = \lceil \log_2(\delta_r) \rceil$.
\end{itemize}

Note that the minimum scale is not the minimum scale of {\it any} cover tree
node (that would be $-\infty$), but the minimum scale of any non-leaf node in
the tree.

Suppose that our datasets are of a similar scale distribution: $s_q^T = s_r^T$,
and $s_q^{\min} = s_r^{\min}$.  In this setting we will have alternating query
and reference recursions.  But if this is not the case, then we have extra
reference recursions before the first query recursion or after the last query
recursion (situations where both these cases happen are possible).  Motivated by
this observation, let us quantify these extra reference recursions:

\begin{lemma}
\label{lem:extcase1}
For a dual-tree algorithm with $S_q \sim S_r \sim O(N)$ using cover trees and
the traversal given in Algorithm \ref{alg:cover-tree-dual}, the number of extra
reference recursions that happen before the first query recursion is bounded by

\begin{equation}
\min\left(O(N), \lceil \log_2(\eta_r / \eta_q) \rceil\right).
\end{equation}
\end{lemma}

\begin{proof}
The first query recursion happens once $s_q > s_r^{\max}$.  The number of
reference recursions before the first query recursion is then bounded as the
number of levels in the reference tree between $s_r^T$ and $s_q^T$ that have at
least one explicit node.  Because there are $O(N)$ nodes in the reference tree,
the number of levels cannot be greater than $O(N)$ and thus the result holds.

The second bound holds by applying the definitions of $s_r^T$ and $s_q^T$ to the
expression $s_r^T - s_q^T$.
\end{proof}

Note that the $O(N)$ bound may be somewhat loose, but it suffices for our later
purposes.  Now let us consider the other case:

\begin{lemma}
\label{lem:extcase3}
For a dual-tree algorithm with $S_q \sim S_r \sim O(N)$ using cover trees and
the traversal given in Algorithm \ref{alg:cover-tree-dual}, the number of extra
reference recursions that happen after the last query recursion is bounded by

\begin{equation}
\theta = \max\left(\min\left(O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
O(N^2)\right), \; 0\right).
\end{equation}
\end{lemma}

\begin{proof}
Our goal here is to count the number of reference recursions after the final
query recursion at level $s_q^{\min}$.  Because query nodes are not pruned in
this traversal, each reference recursion we are counting will be duplicated over
the whole set of $O(N)$ query nodes.  The first part of the bound follows by
observing that $s_q^{\min} - s_r^{\min} = \lceil \log_2(\delta_q) \rceil -
\lceil \log_2(\delta_r) \rceil \le \lceil \log_2(\delta_q / \delta_r) \rceil$.
The second part follows by simply observing that there are $O(N)$ reference
nodes.
\end{proof}

These two previous lemmas allow us a better understanding of what happens as the
reference set and query set become different.  Using them, we will prove general
runtime bounds for the cover tree traversal.

@@

\begin{equation}
O\left(c_r^4 | R^* | \chi \psi (N + i_t(\mathscr{T}_q) + \theta)\right)
\end{equation}

\noindent time, where $ | R^* | $ is the maximum size of the reference set $R$
(line \ref{alg:line:ct-dual-input}) during the dual-tree recursion, $\chi$ is
the maximum possible runtime of \texttt{BaseCase()}, $\psi$ is the maximum
possible runtime of \texttt{Score()}, and $\theta$ is defined as in Lemma
\ref{lem:extcase3}.
\end{thm}

@@

First, consider case \textit{(1)}.  Lemma \ref{lem:extcase1} shows that the
number of reference recursions of this type is bounded by $O(N)$.  Although
there is also a bound that depends on the sizes of the datasets, we only aim to
show a linear runtime bound, so the $O(N)$ bound is sufficient here.

@@

Lastly, for case \textit{(3)}, we may refer to Lemma \ref{lem:extcase3}, giving
a bound of $\theta$ reference recursions in this case.

We may now combine these results for the runtime of a query recursions with the
total number of reference recursions in order to give the result of the theorem:

\begin{equation}
O(c_r^4 \psi \chi | R^* | (N + i_t(\mathscr{T}_q) + \theta)) + O(c_r^4 \psi | R^* | N) \sim O(c_r^4 \psi
\chi |R^*| (N + i_t(\mathscr{T}_q + \theta)).
\end{equation}
\end{proof}

@@

The bichromatic case ($S_q \ne S_r$) is a slightly more complex result which
deserves a bit more attention.  The intuition for all terms except $\theta$
remain virtually the same.

The term $\theta$ captures the effect of query and reference datasets with
different widths, and has one unfortunate corner case: when $\delta_q > \eta_r$,
then the query tree must be entirely descended before any reference recursion.
This results in a bound of the form $O(N \log (\eta_r / \delta_r))$, or
$O(N^2)$.  This is because the reference tree must be descended individually for
each query point.

@@

Our results are similar to that of \citet{ram2009}, but those results depend on
a quantity called the {\it constant of bichromaticity}, denoted $\kappa$, which
has unclear relation to cover tree imbalance.  The dependence on $\kappa$ is
given as $c_q^{4 \kappa}$, which is not a good bound, especially because
$\kappa$ may easily be very large.

@@

Both $\nu$ and $\kappa$ are difficult to empirically calculate and require an
entire run of the dual-tree algorithm.  On the other hand, bounding
$i_t(\mathscr{T}_q)$ (and $\theta$) can be done in one pass of the tree
(assuming the tree is already built).  Thus, not only is our bound tighter when
the cover tree imbalance is sublinear in $N$, it more closely reflects the
actual behavior of dual-tree algorithms, and the constants which it depends upon
are straightforward to calculate.

### 2014-12-24 bounds-files/imbalance.tex
Ready for review.

An imbalanced cover tree can---and often does---happen in practice, and the
imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
Consider a dataset with a single outlier which is very far away from all of the
other points\footnote{Note also that for an outlier sufficiently far away, the
expansion constant is $N - 1$.}.  Figure \ref{fig:outlier} shows what happens in
this situation: the root node has two children; one of these children has only
the outlier as a descendant, and the other child has the rest of the points in
the dataset as a descendant.  In fact, it is easy to find datasets with multiple
outliers which give rise to a chain-like structure at the top of the tree: see
Figure \ref{fig:outliers} for an illustration\footnote{As a side note, this
behavior is not limited to cover trees, and can happen to mean-split $kd$-trees
too, especially in higher dimensions.}.

\begin{figure}
\begin{center}
  \input{figures/single-outlier-tree.tex}
\end{center}
\label{fig:outlier}
\caption{Single-outlier cover tree.}
\end{figure}

@@

\begin{defn}
The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

@@

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that the imbalance is actually quadratic in $N$!

As we will see later empirically, the notion of cover tree imbalance does a good
job of capturing the `goodness' of a tree.  We will use this notion in order to
quantify the worst-case performance of dual-tree algorithms using the cover
tree.

{\bf TODO: Can any bound be proven on $i_t(\mathscr{T})$?  I haven't thought of
anything successfully yet.}

>>>>

An imbalanced cover tree can happen in practice, and in the worst cases, the
imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
Consider a dataset with a single outlier which is very far away from all of the
other points\footnote{Note also that for an outlier sufficiently far away, the
expansion constant is $N - 1$.}.  Figure \ref{fig:outlier} shows what happens in
this situation: the root node has two children; one of these children has only
the outlier as a descendant, and the other child has the rest of the points in
the dataset as a descendant.  In fact, it is easy to find datasets with a
handful of outliers that give rise to a chain-like structure at the top of the
tree: see Figure \ref{fig:outliers} for an illustration\footnote{As a side note,
this behavior is not limited to cover trees, and can happen to mean-split
$kd$-trees too, especially in higher dimensions.}.

\begin{figure}
\begin{center}
  \input{figures/single-outlier-tree.tex}
\end{center}
\caption{Single-outlier cover tree.}
\label{fig:outlier}
\end{figure}

@@

\begin{defn}
\label{def:imbalance}
The {\it cover tree imbalance} $i_t(\mathscr{T})$ for a cover tree $\mathscr{T}$
is defined as the cumulative number of missing levels in the tree.  This can be
expressed as a function of cover node imbalances easily:

@@

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$.

However, for most real-world datasets with the cover tree implementation in {\bf
mlpack} \citep{mlpack2013} and the reference implementation
\citep{langford2006}, the tree imbalance is near-linear with the number of
points.  Most of the cover tree imbalance is contributed by leaf nodes whose
parent has scale greater than $s_{\min}$.  At this time, no cover tree
construction algorithm specifically aims to minimize imbalance.

%To demonstrate this, Table \ref{tab:emp_imbalance} shows empirically
%calculated cover tree imbalance for cover trees on a number of datasets built
%with both the cover tree implementation in {\bf mlpack} \citep{mlpack2013} and
%the original reference implementation \citep{langford2006}.

%\begin{table}[tb]
%\begin{center}
%\begin{tabular}{|c|c|c|r|r|}
%\hline
%Dataset & $N$ & $d$ & reference & mlpack \\
%\hline
%wine & & & & \\
%isolet & & & & \\
%corel & & & & \\
%mnist & 70000 & 784 & & \\
%bio & & & & \\
%phy & & & & \\
%covertype & & & & \\
%LCDM & & & & \\
%\hline
%\end{tabular}
%\end{center}
%\caption{Empirical tree imbalance calculations.}
%\label{tab:emp_imbalance}
%\end{table}

%{\bf TODO: Can any bound be proven on $i_t(\mathscr{T})$?  I haven't thought of
%anything successfully yet.}

### 2014-12-24 bounds-files/sec01-supporting-lemmas.tex
Ready for review.

Note that the cover tree may be interpreted as an infinite-leveled tree, with
$C_{\infty}$ containing only the root point, $C_{-\infty} = S$, and all levels
between defined according to the definitions above.  \citep{langford2006} find
this representation (which they call the {\it implicit} representation) easier
for description of their algorithms and some of their proofs.  But naturally,
this is not suitable for implementation; hence, there is an {\it explicit}
representation in which all nodes that have only a self-child are coalesced.
Figure \ref{fig:implicit-explicit} shows an example cover tree and highlights
implicit and explicit nodes.

@@

Take $f(x)$ as a uniform spherical distribution in $\mathcal{R}^d$: for any $|x|
\le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x) = 0$.  It is easy to see that
$c_f$ in this situation is $2^d$.  To demonstrate this convergence, Table
\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
expansion constants for datasets drawn from this distribution, for varying $d$.

\begin{table}[htb]
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
\hline
$|S|$ & $d = 2$ & $d = 5$ & $d = 10$ & $d = 20$ \\
\hline
10      & 6     & 10    & 10    &         \\
31      & 12    & 31    & 31    &         \\
100     & 13    & 64    & 100   &         \\
316     & 23    & 96    & 286   &         \\
1000    & 26    & 126   & 673   &         \\
3162    & 29    & 189   & 1348  &         \\
10000   & 30    & 259   & 2646  &         \\
31622   & 39    & 309   & 3858  &         \\
100000  & 40    & 360   & 7162  &         \\
316227  & 46    & 385   & 8297  &         \\
1000000 &       &       &       &         \\
$\infty$ ($c_f$)   & 4     & 32    & 1024  & 1048576 \\
\hline
\end{tabular}
\end{center}
\caption{Calculated $c$ for uniform hypersphere distribution as
$|S|$ scales.}
\label{tab:ec_scaling}
\end{table}

We are somewhat limited in the sizes of $|S|$ we can scale to, because
calculating the expansion constant of a dataset is a computationally intensive
task.  Nonetheless, we can see that the expansion constant moves towards the
final value of $c_f$, though potentially from above ({\it The results are not
done yet.  This paragraph may need rewriting.  But I strongly suspect the
results will show what I've said.}).  We postulate that the behavior of more
complex real-world datasets is similar.

>>>>

Note that the cover tree may be interpreted as an infinite-leveled tree, with
$C_{\infty}$ containing only the root point, $C_{-\infty} = S$, and all levels
between defined according to the definitions above.  \citep{langford2006} find
this representation (which they call the {\it implicit} representation) easier
for description of their algorithms and some of their proofs.  But naturally,
this is not suitable for implementation; hence, there is an {\it explicit}
representation in which all nodes that have only a self-child are coalesced.

@@

Take $f(x)$ as a uniform spherical distribution in $\mathcal{R}^d$: for any $|x|
\le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x) = 0$.  It is easy to see that
$c_f$ in this situation is $2^d$. % To demonstrate this convergence, Table
%\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
%expansion constants for datasets drawn from this distribution, for varying $d$.

%\begin{table}[htb]
%\begin{center}
%\begin{tabular}{|l|c|c|c|c|}
%\hline
%$|S|$ & $d = 2$ & $d = 5$ & $d = 10$ & $d = 20$ \\
%\hline
%10      & 6     & 10    & 10    &         \\
%31      & 12    & 31    & 31    &         \\
%100     & 13    & 64    & 100   &         \\
%316     & 23    & 96    & 286   &         \\
%1000    & 26    & 126   & 673   &         \\
%3162    & 29    & 189   & 1348  &         \\
%10000   & 30    & 259   & 2646  &         \\
%31622   & 39    & 309   & 3858  &         \\
%100000  & 40    & 360   & 7162  &         \\
%316227  & 46    & 385   & 8297  &         \\
%1000000 &       &       &       &         \\
%$\infty$ ($c_f$)   & 4     & 32    & 1024  & 1048576 \\
%\hline
%\end{tabular}
%\end{center}
%\caption{Calculated $c$ for uniform hypersphere distribution as
%$|S|$ scales.}
%\label{tab:ec_scaling}
%\end{table}

%We are somewhat limited in the sizes of $|S|$ we can scale to, because
%calculating the expansion constant of a dataset is a computationally intensive
%task.  Nonetheless, we can see that the expansion constant moves towards the
%final value of $c_f$, though potentially from above ({\it The results are not
%done yet.  This paragraph may need rewriting.  But I strongly suspect the
%results will show what I've said.}).  We postulate that the behavior of more
%complex real-world datasets is similar.
